{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-08-03 ident_manu.Rnw
intro draft done

The development and application of a model represents a tradeoff between desirable characteristics that are defined by the analysis needs.  An idealized model should be sufficiently generalizable across systems, provide results that are precise given the inputs, and include structural components that are realistic descriptions of actual processes \citep{Levins66}. Given that a model cannot possibly excel in all of these characterics, models are developed in partial dependence of reality and theoretical constructs, completely separate from both, or dependent on one or the other \citep{Morrison99}.  These challenges are analagous to the well-known `bias-variance' tradeoff in statistical models that balances the competing objectives of over- and under-fitting to an observed dataset. Process-based models are typically not well balanced between reality and theory, such that a common tendency during development is to over-parameterize equations in an attempt to descibe reality completely.  Such over-parameterization, including use of excessive structural equations, can have serious implications for practical applications of a model.  Over-parameterization can limit application to novel systems outside of the development area and impose uncertainty in model predictions as `realistic' values for every variable may not be known.    Quantitative limitations of over-parameterization are also analagous to degrees of freedom in standard statistical models as free parameters cannot be numerically estimated when constrained to an observed dataset.  The application of process-based models to describe hypoxia dynamics has not been immune to these challenges and more comprehensive approaches are needed to develop models that more carefully balance theory with reality.

Standard approaches for uncertainty analysis can be used

\begin{comment}
\begin{enumerate}
\item What models seek to provide - generality, precision, realism \cite{Levins66}, there is a tradeoff so models are 1) developed in partial independence and dependence on the world and theory, 2) function autonomously from both, or 3) represent both at the same time, from \cite{Morrison99}, cited in \cite{Ganju16}.  This is similar to the bias-variance tradeoff for statistical models, e.g., overparameterization of a model makes it very biased as it fits the data (the world) exactly, tradeoff between sensitity and error with changes in model complexity (more complexity is less error but increasing sensitivity) described in \cite{Snowling01}
\item How is model performance/uncertainty evaluated regarding what they should provide - structural, observational, parameter \cite{Beck87}? \cite{Refsgaard07} provides comprehensive overview of factors that contribute to uncertainty in environmental modelling.
\item Parameter uncertainy as low-hanging fruit - can do post-hoc and from inner to outer level of complexity, parameter uncertainty is the most common, e.g., marine ecological model \cite{Mateus15}, but stopped short, global sensitivity analysis of eutrophication model \cite{Estrada10}
\item Challenges related to uncertainty - similar to degrees of freedom, identifiability definition from \cite{Brun01} and need to evaluate identifiability \cite{Fasham06}, \cite{Omlin01} did a similar analysis with freshwater biogeochem model.
Identifiability describes the ability to estimate a parameter in relation to variation among the remaining parameters.  A parameter is identifiable if all parameters within the set can be uniquely estimated based on the observed data.  Parameters that are unidentifiable typically produce similar model outputs for a given relative perturbation, i.e., the effect of altering one parameter can be undone by altering one or more other parameters.  Model calibration will not converge for parameters sets that are unidentifiable. 
\end{enumerate}
\end{comment}

@@

@@

A final point about optimization with identifiable parameter sets - optimization to fit the data still does not ensure a correct model.  Failing in one way can be over-compensated by another feature, e.g., the parameter set that is optimized (see \cite{Flynn05}, p. 1207, third paragraph), also \citep{Arhonditsis08}, also note that over-parameterized models are not necessarily bad, see \citep{Omlin01}

@@

Our heuristic products are partially analogous to the Rainfall-Runoff Modelling Toolbox (RRMT) presented by Wagener et al. 1999, 2001 (cited on page 15, in \citealt{Wagener01}).

@@

>>>>

The development and application of a model represents a tradeoff between different characteristics that are expected from the output or provided by the structural components. An idealized model is sufficiently generalizable across systems, provides results that are precise given the inputs, and includes components that are realistic descriptions of actual processes \citep{Levins66}. Given that all of these characteristics cannot be achieved, models are developed in partial dependence of reality and theoretical constructs, completely separate from both, or dependent on one or the other \citep{Morrison99,Ganju16}.  These challenges are analagous to the well-known `bias-variance' tradeoff in statistical models that balances the competing objectives of over- and under-fitting an observed dataset. Process-based models are more commonly imbalanced between reality and theory, such that many models are over-parameterized in an attempt to describe reality completely \citep{Denman03,Nossent12,Petrucci14}.  Such over-parameterization, including use of excessive structural equations, can have serious implications for practical applications of a model.  Over-parameterization can limit application to novel systems outside of the data domain and impose uncertainty in model predictions as `realistic' values for every variable may not be known or inaccurately applied from existing studies \citep{Durand02,Refsgaard07,Wade08}. More importantly, quantitative limitations of over-parameterization are analagous to degrees of freedom in standard statistical models as free parameters cannot be numerically estimated when constrained to an observed dataset \citep{Kirchner06}.  The application of process-based models to describe hypoxia dynamics has not been immune to these challenges and more comprehensive approaches are needed to develop models that more carefully balance theory with reality \citep[e.g.,][]{Snowling01}.

Standard approaches for uncertainty analysis can be used to begin addressing model complexity issues.  In the most general sense, uncertainty can be evaluated relative to the effects of input conditions or the observed data used to calibrate a model, changes in parameter values, or variation in the structural components \citep{Beck87}.  Evaluating parameter uncertainty is by far the most common and simplest means of evaluating model behavior.  Although uncertainty analyses should be integrated throughout model development and application, parameters are more often evaluated post-hoc as a form of `damage control' for further calibration.  This approach is related to inverse modelling where results from sensitivity analysis of parameters are used to guide calibration or fit of the developed model to observations (\citealt{Soetaert10}, or confronting models with data, \textit{sensu} \citealt{Hilborn97}).  Parameter sensitivity analysis combined with inverse modelling necessarily involves questions of parameter `identifiability', such that only a subset of parameters can be numerically constrained to the data as compared to the entire parameter set.  Redundancies in the effects of parameters on the output lead to unidentifiable models where optimal solutions may be empirically impossible (i.e., standard algorithms will not converge) or parameter values may be non-unique leading to the right answer for the wrong reason \citep{Kirchner06}.  The concept of identifiable parameter subsets is not foreign to hypoxia or eutrophication models  \citep{Omlin01,Estrada10,Mateus15}, although there is a clear need for greater integration of these concepts in model development \citep{Fasham06}.  Moreover, the inclusion of sensitivity and identifiability analyses in model tuning will require the adoption of selection rules that are context-dependent given the subsets that are possible with large parameter sets \citep[e.g.,][]{Wagener01, Wagener01b}.

\begin{comment}
\begin{enumerate}
\
\item Parameter uncertainy as low-hanging fruit - can do post-hoc and from inner to outer level of complexity, parameter uncertainty is the most common, e.g., marine ecological model \cite{Mateus15}, but stopped short, global sensitivity analysis of eutrophication model \cite{Estrada10}
\item Challenges related to uncertainty - similar to degrees of freedom, identifiability definition from \cite{Brun01} and need to evaluate identifiability \cite{Fasham06}, \cite{Omlin01} did a similar analysis with freshwater biogeochem model.
Identifiability describes the ability to estimate a parameter in relation to variation among the remaining parameters.  A parameter is identifiable if all parameters within the set can be uniquely estimated based on the observed data.  Parameters that are unidentifiable typically produce similar model outputs for a given relative perturbation, i.e., the effect of altering one parameter can be undone by altering one or more other parameters.  Model calibration will not converge for parameters sets that are unidentifiable. 
\end{enumerate}
\end{comment}

@@

\citep{Denman03} describes over-parameterization for multiple phyto groups, use this to emphasize collinearity issues with the phyto groups despite differing sensitivity values between groups

@@

A final point about optimization with identifiable parameter sets - optimization to fit the data still does not ensure a correct model.  Failing in one way can be over-compensated by another feature, e.g., the parameter set that is optimized (see \cite{Flynn05}, p. 1207, third paragraph), also \citep{Durand02,Arhonditsis08}, also note that over-parameterized models are not necessarily bad, see \citep{Omlin01}, models must be validated for the uses which they were intended where a comparison of observed and predicted is the simplest approach and other methods should be used depending on analysis needs \citep{Rykiel96}

@@

Our heuristic products are partially analogous to the Rainfall-Runoff Modelling Toolbox (RRMT) presented by \citep{Wagener01b}2001 (cited on page 15, in \citealt{Wagener01}), although ours is simple in comparison

@@

  }) %>% 
  do.call('rbind', .) %>%
  mutate(
    error = sapply(error, scinot),
    coll = sapply(coll, scinot), 
    Rank = sapply(rnk_incat, scinot), 
    Rank = paste0(Rank, '\\textsuperscript{', Category, '}')
    ) %>%
  group_by(stts) %>% 
  nest %>% 
  unnest %>% 
  select(stts, lngs, Parameter, Rank, error, coll) %>% 
  rename(
    Description = lngs, 
    L1 = error
  ) %>% 
  mutate(
    stts = as.character(stts),
    Description = paste0('\\footnotesize{', Description, '}'),
    Parameter = paste0('\\footnotesize{', Parameter, '}')
    ) %>% 
  arrange(stts)

### 2016-08-03 ident_manu.tex
intro draft done

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Tue Aug 2 11:40:44 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

% ident all states
%latex.default(identall1, file = "", rowlabel = "Parameter description",     caption = cap_val, caption.loc = "top", size = "normalsize",     label = "tab:identall1", rowname = descrip, colheads = colnms,     rgroupTexCmd = NULL)%
\begin{table}[!tbp]
{\normalsize
\caption{Parameter identifiability for dissolved oxygen.  Selection followed the second heuristic where parameters were selected by decreasing sensitivity up to maximum $\gamma < 15$.  Rank describes the parameter sensitivity in each category (O: optics, OM: organic matter, P: phytoplankton, T: temperature, Z: zooplankton).\label{tab:identall1}} 
\begin{center}
\begin{tabular}{lllll}
\hline\hline
\multicolumn{1}{l}{Parameter description}&\multicolumn{1}{c}{Parameter}&\multicolumn{1}{c}{Rank}&\multicolumn{1}{c}{L1}&\multicolumn{1}{c}{$\gamma$}\tabularnewline
\hline
\scriptsize{phytoplankton carbon/cell}&\scriptsize{\textit{Qc$_{p1}$}}&$1$\textsuperscript{P}&$328.35$&$1$\tabularnewline
\scriptsize{phytoplankton carbon/cell}&\scriptsize{\textit{Qc$_{p3}$}}&$2$\textsuperscript{P}&$232.8$&$1$\tabularnewline
\scriptsize{N-uptake rate measured at umax}&\scriptsize{\textit{vmaxN$_{p5}$}}&$3$\textsuperscript{P}&$7.57$&$1$\tabularnewline
\scriptsize{phytoplankton carbon/cell}&\scriptsize{\textit{Qc$_{p2}$}}&$4$\textsuperscript{P}&$4.08$&$1$\tabularnewline
\scriptsize{phytoplankton basal respiration coefficient}&\scriptsize{\textit{respb$_{p4}$}}&$5$\textsuperscript{P}&$3.1$&$1$\tabularnewline
\scriptsize{mortality coefficient}&\scriptsize{\textit{mA$_{p3}$}}&$6$\textsuperscript{P}&$2.98$&$1.01$\tabularnewline
\scriptsize{Tref(nospA+nospZ): Optimum temperature for growth(C)}&\scriptsize{\textit{Tref(nospA+nospZ)$_{p4}$}}&$1$\textsuperscript{T}&$2.9$&$1.02$\tabularnewline
\scriptsize{N-uptake rate measured at umax}&\scriptsize{\textit{vmaxN$_{p4}$}}&$7$\textsuperscript{P}&$2.88$&$1.04$\tabularnewline
\scriptsize{coefficient for non-limiting nutrient}&\scriptsize{\textit{aN$_{p1}$}}&$8$\textsuperscript{P}&$2.7$&$1.05$\tabularnewline
\scriptsize{P-uptake rate measured at umax}&\scriptsize{\textit{vmaxP$_{p3}$}}&$9$\textsuperscript{P}&$2.23$&$1.07$\tabularnewline
\scriptsize{phytoplankton basal respiration coefficient}&\scriptsize{\textit{respb$_{p3}$}}&$10$\textsuperscript{P}&$2.15$&$1.73$\tabularnewline
\scriptsize{rate constant for nitrification}&\scriptsize{\textit{k11}}&$1$\textsuperscript{OM}&$2.11$&$1.73$\tabularnewline
\scriptsize{N-uptake rate measured at umax}&\scriptsize{\textit{vmaxN$_{p3}$}}&$11$\textsuperscript{P}&$1.99$&$1.73$\tabularnewline
\scriptsize{O2 concentration that inhibits denitrification}&\scriptsize{\textit{KstarO2}}&$2$\textsuperscript{OM}&$1.91$&$1.73$\tabularnewline
\scriptsize{Chla specific absorption at 490 nm}&\scriptsize{\textit{astar490}}&$1$\textsuperscript{O}&$1.64$&$1.77$\tabularnewline
\scriptsize{initial slope of the photosynthesis-irradiance relationship}&\scriptsize{\textit{alpha$_{p4}$}}&$12$\textsuperscript{P}&$1.61$&$4.1$\tabularnewline
\scriptsize{minimum N cell-quota}&\scriptsize{\textit{QminN$_{p3}$}}&$13$\textsuperscript{P}&$1.53$&$4.1$\tabularnewline
\scriptsize{turnover rate for OM1A and OM1G}&\scriptsize{\textit{KG1}}&$3$\textsuperscript{OM}&$1.49$&$4.2$\tabularnewline
\scriptsize{turnover rate for OM2A and OM2G}&\scriptsize{\textit{KG2}}&$4$\textsuperscript{OM}&$1.37$&$4.62$\tabularnewline
\scriptsize{mortality coefficient}&\scriptsize{\textit{mA$_{p4}$}}&$14$\textsuperscript{P}&$1.35$&$4.62$\tabularnewline
\hline
\end{tabular}\end{center}}

>>>>

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Tue Aug 2 17:30:02 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

% ident all states
%latex.default(identall1, file = "", rowlabel = "Parameter description",     caption = cap_val, caption.loc = "top", size = "normalsize",     label = "tab:identall1", rowname = descrip, colheads = colnms,     rgroupTexCmd = NULL)%
\begin{table}[!tbp]
{\normalsize
\caption{Parameter identifiability for dissolved oxygen.  Selection followed the second heuristic where parameters were selected by decreasing sensitivity up to maximum $\gamma < 15$.  Rank describes the parameter sensitivity in each category (O: optics, OM: organic matter, P: phytoplankton, T: temperature, Z: zooplankton).\label{tab:identall1}} 
\begin{center}
\begin{tabular}{lllll}
\hline\hline
\multicolumn{1}{l}{Parameter description}&\multicolumn{1}{c}{Parameter}&\multicolumn{1}{c}{Rank}&\multicolumn{1}{c}{L1}&\multicolumn{1}{c}{$\gamma$}\tabularnewline
\hline
\footnotesize{phytoplankton carbon/cell}&\footnotesize{\textit{Qc$_{p1}$}}&$1$\textsuperscript{P}&$328.35$&$1$\tabularnewline
\footnotesize{phytoplankton carbon/cell}&\footnotesize{\textit{Qc$_{p3}$}}&$2$\textsuperscript{P}&$232.8$&$1$\tabularnewline
\footnotesize{N-uptake rate measured at umax}&\footnotesize{\textit{vmaxN$_{p5}$}}&$3$\textsuperscript{P}&$7.57$&$1$\tabularnewline
\footnotesize{phytoplankton carbon/cell}&\footnotesize{\textit{Qc$_{p2}$}}&$4$\textsuperscript{P}&$4.08$&$1$\tabularnewline
\footnotesize{phytoplankton basal respiration coefficient}&\footnotesize{\textit{respb$_{p4}$}}&$5$\textsuperscript{P}&$3.1$&$1$\tabularnewline
\footnotesize{mortality coefficient}&\footnotesize{\textit{mA$_{p3}$}}&$6$\textsuperscript{P}&$2.98$&$1.01$\tabularnewline
\footnotesize{Tref(nospA+nospZ): Optimum temperature for growth(C)}&\footnotesize{\textit{Tref(nospA+nospZ)$_{p4}$}}&$1$\textsuperscript{T}&$2.9$&$1.02$\tabularnewline
\footnotesize{N-uptake rate measured at umax}&\footnotesize{\textit{vmaxN$_{p4}$}}&$7$\textsuperscript{P}&$2.88$&$1.04$\tabularnewline
\footnotesize{coefficient for non-limiting nutrient}&\footnotesize{\textit{aN$_{p1}$}}&$8$\textsuperscript{P}&$2.7$&$1.05$\tabularnewline
\footnotesize{P-uptake rate measured at umax}&\footnotesize{\textit{vmaxP$_{p3}$}}&$9$\textsuperscript{P}&$2.23$&$1.07$\tabularnewline
\footnotesize{phytoplankton basal respiration coefficient}&\footnotesize{\textit{respb$_{p3}$}}&$10$\textsuperscript{P}&$2.15$&$1.73$\tabularnewline
\footnotesize{rate constant for nitrification}&\footnotesize{\textit{k11}}&$1$\textsuperscript{OM}&$2.11$&$1.73$\tabularnewline
\footnotesize{N-uptake rate measured at umax}&\footnotesize{\textit{vmaxN$_{p3}$}}&$11$\textsuperscript{P}&$1.99$&$1.73$\tabularnewline
\footnotesize{O2 concentration that inhibits denitrification}&\footnotesize{\textit{KstarO2}}&$2$\textsuperscript{OM}&$1.91$&$1.73$\tabularnewline
\footnotesize{Chla specific absorption at 490 nm}&\footnotesize{\textit{astar490}}&$1$\textsuperscript{O}&$1.64$&$1.77$\tabularnewline
\footnotesize{initial slope of the photosynthesis-irradiance relationship}&\footnotesize{\textit{alpha$_{p4}$}}&$12$\textsuperscript{P}&$1.61$&$4.1$\tabularnewline
\footnotesize{minimum N cell-quota}&\footnotesize{\textit{QminN$_{p3}$}}&$13$\textsuperscript{P}&$1.53$&$4.1$\tabularnewline
\footnotesize{turnover rate for OM1A and OM1G}&\footnotesize{\textit{KG1}}&$3$\textsuperscript{OM}&$1.49$&$4.2$\tabularnewline
\footnotesize{turnover rate for OM2A and OM2G}&\footnotesize{\textit{KG2}}&$4$\textsuperscript{OM}&$1.37$&$4.62$\tabularnewline
\footnotesize{mortality coefficient}&\footnotesize{\textit{mA$_{p4}$}}&$14$\textsuperscript{P}&$1.35$&$4.62$\tabularnewline
\hline
\end{tabular}\end{center}}
